### 2021-06-12 abstractESSLLI21/abstractESSLLI1.tex
Adds models names

\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as number vectors (see e.g. {[}1{]} and {[}4{]} and
references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

@@

Such methods are statistically problematic. One issue, specific to
{[}6{]}, is that no distinction is made based on whether given a
protected word and a class of attributes is stereotypically associated
with this protected word or a different protected word. A related,
general problem, is that all the authors ignore the step of comparing
their results with control groups, especially control groups of
stereotype-neutral human attributes.
\todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"? albo jakoś podobnie}
That such a move is important is suggested for instance by Figure 1
prepared using the original word list for religion-related protected
words extended with control attributes, where such human attributes are
also closer to protected classes than 1.

Another, also serious problem, is that all the authors calculate means
of means and their authors
\todo{jak rozumiec drugi raz powtorzone their authors?} run statistical
tests on sets of means. Unfortunately, by pre-averaging the data they
throw away information about sample sizes, and they remove variation,
and so pre-averaging tends to manufacture false confidence.

All such approaches come up with rather short lists of protected words
and rather short lists of stereotypical attributes. Clearly, these are
not complete list. So let's treat them as samples from richer pools of
stereotypical predicates and let's take the uncertainty and variation
involved seriously. To illustrate, let's employ the formulas used by
{[}2{]} in a simple simulation.
\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}

@@

We build analogous models for other topic groups (race, gender), and
different embeddings (Google, and hard-debiased Reddit). The general
conclusion is that once we proceed this way, the situation is much less
obvious. The word list sizes are small, sample sizes are small, and so
posterior density intervals are wide. Some stronger bias can be
discerned in the gender case. Moreover, sometimes the differences
between associated, different and human, are not very impressive. The
cost of debiasing turns out sometimes to be that neutral human
predicates get closer to protected classes (in religion), change in
gender after debiasing is really minor with zero still out out of HPDI
range, and debiasing improvement for race is a small improvement
achieved at the price of moving neutral terms closer to protected words
(there is no space to visualise these results in this abstract).

>>>>

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
and references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

@@

Such methods are statistically problematic. Two issues, specific to
{[}6{]}, are that (i) no distinction is made based on whether given a
protected word and a class of attributes is stereotypically associated
with this protected word or a different protected word, and (ii) no
re-scaling by standard deviation (typical in effect size measures such
as Cohen's \(d\)) is used, so MAC is not a good measure of effect size.
A more general problem, is that all the authors ignore the step of
comparing their results with control groups, especially control groups
of stereotype-neutral human attributes. That such a move is important is
suggested for instance by Figure 1 prepared using the original word list
for religion-related protected words extended with control attributes,
where such human attributes are also closer to protected classes than
neutral non-human words.

Another, also serious problem, is that all the authors calculate means
of means and run statistical tests on sets of means. Unfortunately, by
pre-averaging the data they throw away information about sample sizes,
and they remove variation, and so pre-averaging tends to manufacture
false confidence.

All such approaches come up with rather short lists of protected words
and rather short lists of stereotypical attributes. Clearly, these are
not complete list. So let's treat them as samples from richer pools of
stereotypical predicates and let's take the uncertainty and variation
involved seriously. To illustrate, let's employ the formulas used by
{[}2{]} in a simple simulation.

@@

We build analogous models for other topic groups (race, gender), and
different embeddings (we compare the results of two models:
GoogleNews-vectors-negative300, and Reddit Corpus). The general
conclusion is that once we proceed this way, the situation is much less
obvious. The word list sizes are small, sample sizes are small, and so
posterior density intervals are wide. Some stronger bias can be
discerned in the gender case. Moreover, sometimes the differences
between associated, different and human, are not very impressive. The
cost of debiasing turns out sometimes to be that neutral human
predicates get closer to protected classes (in religion), change in
gender after debiasing is really minor with zero still out out of HPDI
range, and debiasing improvement for race is a small improvement
achieved at the price of moving neutral terms closer to protected words
(there is no space to visualise these results in this abstract).
